Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2021-10-18 numerics_paper/version2/Response_to_Reviewers_R2.tex

\documentclass[english,11pt]{article}

\usepackage{enumitem}
\usepackage{cite}
\usepackage{bbm}
\usepackage{ulem}
\usepackage{amsthm}
\usepackage[latin9]{inputenc}
\usepackage{babel}
\usepackage[hmargin=0.9in, vmargin=0.75in]{geometry}

\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage[dvipsnames]{xcolor}

\usepackage{url}
\usepackage{bm}

\usepackage{amsmath}
\usepackage[numbers]{natbib}
%\let\cite=\citet
\newcommand{\bfx}{{\bf x}}

\makeatletter
\providecommand{\tabularnewline}{\\}
\makeatother

%%---------------------------------------------------%%
%%---------------------- TITLE ----------------------%%
%%---------------------------------------------------%%
\begin{document}
\title{Response to Reviewers}
\maketitle

\vspace{1.5cm}
\noindent
Michael Dumbser \\
Editor \\
International Journal for Numerical Methods in Fluids \\

\vspace{1.5cm}

\noindent
Manuscript \# FLD-21-0079.R1
\vspace{1.5cm}

\noindent
Dear Prof. Michael Dumbser,

\vspace{0.25cm}
\bigskip
The manuscript entitled
\begin{center}
``Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump''
\end{center}
by David I. Ketcheson and Manuel Quezada de Luna
has been revised following the comments made by the referees.

\bigskip
Letters detailing how the comments of each reviewer have been
addressed are enclosed. To facilitate your perusal of the revised
manuscript, changes concerning the comments of the first and second
reviewers are highlighted in red and green, respectively.

\vspace{0.25cm}
\bigskip\noindent 
With our best regards,

\vspace{0.5cm}
\bigskip\noindent
David I. Ketcheson and Manuel Quezada de Luna

\newpage
%%---------------------------------------------------------%%
%%-------------------- FIRST REVIEWER -------------------- %%
%%---------------------------------------------------------%%
\section*{Revision of Manuscript \# FLD-21-0079.R1 -- Reviewer $\#1$}
We thank the reviewer for the comments and suggestions.
We have incorporated all of them.  For your convenience,
the modifications related to the comments made by reviewer $\#1$ and $\#2$
appear in red and green respectively in the new version of the manuscript. 
Statements in red below are comments from the reviewer.
In this letter, the comments made by the reviewer refer to the original
manuscript and our responses refer to the new manuscript.

{\color{red}
  The authors have closely applied the reviewer's suggestions 
  and the quality of the paper has been improved. This manuscript contains 
  a highly valuable work, very relevant to the journal, which may spur 
  some new ideas in the field of numerical shockwave anomalies. 
  Therefore, I do recommend this paper for publication after a minor correction (see below).

  Before publication, I would ask the authors to remove the sentence in lines 14-15. 
  As far as I understand, the slowly moving shock anomaly does not necessarily 
  involve the oscillation of a steady shock. It also involves the presence of the spike. 
  When the shock is not steady and moves slowly, the spike swings and sheds oscillations 
  downstream. However, when the shock is steady, the spike may remain fixed.
}

\noindent
%
%Following the reviewer's suggestion, we have removed the sentence in lines 14-15.

%%----------------------------------------------------------%%
%%-------------------- SECOND REVIEWER -------------------- %%
%%----------------------------------------------------------%%
\newpage
\section*{Revision of Manuscript \# FLD-21-0079.R1 -- Reviewer $\#2$}
We thank the reviewer for the comments and suggestions.
We have incorporated all of them.  For your convenience,
the modifications related to the comments made by reviewer $\#1$ and $\#2$
appear in red and green respectively in the new version of the manuscript.
Statements in green below are comments from the reviewer.
In this letter, the comments made by the reviewer refer to the original
manuscript and our responses refer to the new manuscript.

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(1)]
    To strengthen the notion of using entropy-control as a means to remove the carbuncle, 
    I strongly suggest to cite the most recent work in removing shock instability using 
    entropy control. Chizari et al.
    (CAMWA 2021) developed a more general (and more multidimensional) entropy-stable finite volume
    scheme which has many benefits including minimal sensitivity to grid skewness and does not 
    suffer from shock-instability as shown for the Quirk's problem (Section 5.3). 
    Perhaps, citation of this paper can be done in page 2 in the authors' paper where 
    references [19,20] are cited.

    Cell-vertex entropy-stable finite volume methods for the system of Euler equations 
    on unstructured grids, CAMWA 2021, Vol 98, pp 261-279.
  \end{itemize}
}

\noindent
Thank you for bringing this reference to our attention. We have included the reference as suggested.

\bigskip
{\color{OliveGreen}
  \begin{itemize}
  \item[(2)]
    For the flow past a cylinder problem, I still believe that the choice grids will be a 
    crucial factor in fully understanding the instability unless one can develop a scheme 
    which is minimally sensitive to grid changes as done by Chizari et al (CAMWA 2021). 
    As shown in that paper, the classic entropy-stable FV scheme will eventually succumb to 
    shock instability if the grids are 'bad' enough but the newly developed entropystable 
    FV method (residual-distribution based) therein does not, since the latter is proven to be 
    almost non-affected by grid degradations.

    In the authors' case. Even on a nice grid like the one used in 4.3, there are some 
    issues with convergence (Figure 4b) using the blended 2nd order scheme where there 
    seem to be limit cycle in the residual convergence. The limit cycle may be an onset 
    for instability if the simulation is ran long enough. This is the classic instability 
    issue that has troubled the carbuncle community for decades. The authors' claim
    that this is due to the second order influence, which is fair enough.

    \begin{itemize}
    \item[(a)]
      However, how do you guarantee that the residual of the second order method will completely decay?
      Fig 4b does not indicate that this will happen. In other words, how do you know if the 
      second-order residual limit cycle will not manifest into shock instability if you ran the 
      simulations longer?      
    \end{itemize}
  \end{itemize}
}
\noindent
For a given mesh, we believe the residual will not decay anymore. 
This is a common problem with many 2nd order schemes that make use of limiters 
based on a posteriori fixes; see for instance \cite{kuzmin2020monolithic}. As stated in the 
paper, we believe the reason for the lack of convergence to a steady state is related the second-order 
correction (including the limiters). This is suggested by the fact that the blended solver with 
the first-order scheme does not prevent convergence to a steady state and since it is well-known that 
some a posteriori limiters inhibit convergence to steady state solutions.

We have run the simulation for a much longer time ($t=16000$ instead of $t=160$) and we still do not 
observe a shock instability. Of course this is not a guarantee that a problem will never occur. 
We have modified the text to discuss this problem.

{\color{OliveGreen}
  \begin{itemize}
    \item[]
    \begin{itemize}
      \item[(b)]
        Looking at the magnitude of residual of $\mathcal{O}(10^{-3})$ for the blended 
        2nd order scheme. This is of the same magnitude with the residuals of the HLLEMCC 
        (Fig 4b). The authors state in page 14 that the HLLEMCC residual stagnates. 
        Is this is a bad thing? If yes, how would it be any different from the similar 
        residual behavior for the second order blended method?
    \end{itemize}
  \end{itemize}
}
\noindent
In the paper we state ``With HLLEMCC or the blended solver, the residual stagnates.'' That is, 
we acknowledge that with both solvers the residual stagnates. In our opinion, the fact that the 
residual stagnates is not necessarily a bad thing, assuming this does not lead to a 
serious shock instability, as remarked by the reviewer. We only have numerical evidence that 
this does not occur. In any case, having a method for which the residual of a steady state solution 
decays to zero (up to machine precision) is desirable. We have started a new line of research to 
fix this problem; however, we believe this new research should be included in a different publication. 
We have modified the text (in Section 4.3 and in the conclusions) to discuss this problem and 
to comment on the new line of work we are pursuing.

{\color{OliveGreen}
  \begin{itemize}
    \item[]
    \begin{itemize}
      Is the blended solution in Fig 3c based on the first or second order method? 
      I suggest to included the first order (Fig 3c?) and second order (Fig 3d?) 
      blended shock profiles to demonstrate the difference between the two on a contour plot.
    \end{itemize}
  \end{itemize}
}
\noindent
The solution is based on the second-order method. As suggested by the reviewer, we have included 
a new figure using the first-order method.

\bibliographystyle{plain}
\bibliography{refs}

\end{document}

### 2021-10-18 numerics_paper/version2/numerical_chj.tex
change beta for C
make first changes related to the 2nd revision

In \cite{peery1988blunt} a numerical instability was observed to
appear near the symmetry plane in the simulation of a bow shock.
This phenomenon, subsequently dubbed ``carbuncle", has been observed by many researchers
in similar numerical experiments for the Euler equations, and many remedies
have been proposed, mainly in the form of additional numerical dissipation
\cite{quirk1997contribution,pandolfi2001numerical,dumbser2004matrix,chauvat2005shock,ismail2009proposed,shen2014stability}.
Most notably, dissipative Riemann solvers like HLLE and Rusanov suppress the carbuncle instability \cite{quirk1997contribution}.
For a recent review of numerical shock instability and work to alleviate it,
we refer to \cite[Section 2.5]{simonnumerical} and \cite{zaide2012numerical}.
{\color{OliveGreen}
  The slowly-moving shock anomaly is another important unphysical behavior that might be present 
  with some numerical schemes. We refer to \cite{zaide2012numerical} for a comprehensive 
  discussion about this phenomenon.
  In \cite{navas2019improved}, the authors focus on this problem when solving the shallow water equations. 
}

Given the similarity of structure between the Euler equations and the shallow
water equations, it is not surprising that carbuncles appear in numerical
solutions of the latter as well \cite{kemm2014note}.
The shallow water carbuncle behaves similarly to the
Euler carbuncle; for instance, it appears when the Roe solver is used, but not when
the HLLE or Rusanov solver is used, and can also be suppressed by particular modifications
of the Roe solver \cite{kemm2014note,bader2014carbuncle}.  Another approach to avoid
shallow water carbuncles has been proposed in \cite{navas2019improved}, where it is
studied together with the slowly-moving shock anomaly and in the presence of varying bathymetry.
In this work, we propose a new Riemann solver that blends those of Rusanov and Roe.
%based on a local measure of the entropy residual, 
%as in \cite{guermond2011entropy,guermond2018second,guermond2018well}.
{\color{red} Since Rusanov's solver does not suffer from carbuncle instabilities,
  our aim is to use it near shocks without degrading the good accuracy of Roe's solver. 
  To do this, we need to find a blending criteria that can robustly and sharply locate shocks.
  The entropy residual has proven to be a successful tool for this task; see for instance
  \cite{guermond2011entropy,guermond2018second,guermond2018well}.
  By doing this, our blended solver induces dissipation of entropy.
  Entropy dissipative Riemann solvers have been used to produce carbuncle free solvers for the Euler equations;
  see for instance \cite{ismail2009affordable,ismail2009proposed}.
  However, to the best of our knowledge, this idea has not been used before to supress carbuncle instabilities for the
  shallow water equations. 
}
Using this Riemann solver (within the second-order Lax-Wendroff-LeVeque 
finite volume scheme \cite{leveque1997wave, leveque2002finite}) supresses the formation of carbuncles 
while maintaining an accuracy similar to that of Roe solver.  
{\color{OliveGreen}
  Based on our numerical experiments, using Rusanov's solver does not lead to the 
  slowly-moving shock anomaly. Our blended solver inherits the same property.
}

@@

\subsection{Rusanov's Riemann solver}\label{sec:rusanov}
Rusanov's solver approximates the Riemann solution with a single wave traveling in each
direction.  Both waves are assumed to travel with speed $\lambda^{\max}_\imh$, which is an
upper bound on the wave speeds appearing in the true solution of the Riemann problem.
In this work, we take $\lambda_{i-1/2}^{\max}$ to be the upper bound
from \cite[Prop. 3.7]{azerad2017well}.
The waves are then given by
\begin{align}\label{rusanov_waves}
  \W_{i-1/2}^{1,\Rus}=\bfq_{i-1/2}-\bfq_{i-1}, \qquad
  \W_{i-1/2}^{2,\Rus}=\bfq_{i}-\bfq_{i-1/2},
\end{align}
where the intermediate state $\bfu_{i-1/2}$ is determined by imposing conservation:
\begin{align}\label{rus_cons}
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{1,\Rus}+\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus} = \bff(\bfq_{i})-\bff(\bfq_{i-1}).
\end{align}
The fluctuations are given by \eqref{fluct} with $s_{i-1/2}^1=-\lambda_{i-1/2}^{\max}$
and $s_{i-1/2}^2=\lambda_{i-1/2}^{\max}$, and the first- and second-order methods based
on Rusanov's solver are given by \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}, respectively.
{\color{red}
  The second-order method can also be written in flux-differencing form \eqref{flux-differencing-form}, where 
  \begin{align}\label{first-order_fluxes_LLF}
    \bfF_{i+1/2} = \frac{\bff(\bfq_{i+1})+\bff(\bfq_i)}{2}
    - \frac{\lambda_{i+1/2}^{\max}}{2}(\bfq_{i+1}-\bfq_{i}).
  \end{align}
}

\subsection{Roe's Riemann solver} \label{sec:roe}
The Roe Riemann solver is based on evaluating the flux Jacobian using Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
The waves and speeds are given by the eigenvectors and eigenvalues of the approximate
flux Jacobian obtained using these averages, resulting in
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}
and $\W_{i-1/2}^{p,\Roe}=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with
\begin{align*}
  \hat \bfr^1_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad
  \hat \bfr^2_{i-1/2} =
  \begin{bmatrix}
    0\\
    0\\
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
and the factors $\alpha^p_\imh$ obtained by solving
\begin{align}\label{system_for_alphas}
  \sum_p\W_{i-1/2}^{p,\Roe} = \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=\bfq_i-\bfq_{i-1}.
\end{align}
{\color{red}
  The first-order Roe's fluxes used in \eqref{flux-differencing-form} and \eqref{first-order_via_fluct} are given as follows:
  \begin{align}\label{first-order_fluxes_Roe}
    \bfF_{i+1/2} = \frac{\bff(\bfq_{i+1})+\bff(\bfq_i)}{2} - \frac{1}{2} |\hat A_{i+1/2}|(\bfq_{i+1}-\bfq_{i}).
  \end{align}
}

@@

{\color{red}

@@

}

@@

\subsection{Dam break problem on a dry bed}\label{sec:rp_dry_bed}
We start with the one-dimensional dam break problem on a dry bed.
This problem is a canonical example that demonstrates the `entropy glitch' of Roe's solver
at transonic rarefactions. Since the baseline Roe solver we use (see \S\ref{sec:roe}) does not contain
an entropy fix, it is important to demonstrate that the blended Riemann solver fixes the entropy glitch.
We follow the setup in \cite[\S 4.1.2]{delestre2013swashes}.
The domain is given by $x\in(0,10)$, and the initial condition is ${\color{cyan}u(x,0)=0}$ and
\begin{align*}
  h(x,0)=
  \begin{cases}
    h_l & x\leq x_0=5 \\
    h_r & x >x_0,
  \end{cases}
\end{align*}
with $h_l=5\times 10^{-3}$.  In our experiments we use $h_r=1\times
10^{-15}$ to avoid division by zero.
{\color{OliveGreen}
If we use Rusanov's solver with the first-order method \eqref{first-order_via_fluct}, 
the water height is guaranteed to remain positive. Using Roe or the blended solvers 
can introduce violations of positivity. Alternatively, if we consider the second-order scheme 
\eqref{second-order_via_fluct}, the water height could become negative regardless of the Riemann solver.
To reduce the occurance of loss of positivity, we only use the first-order method 
\eqref{first-order_via_fluct} for this test problem. 
Although using Roe or the blended solvers might still lead to loss of positivity, we 
did not observe that in our experiments.}
The exact solution, which can also be found in \cite{delestre2013swashes} and references therein, is
\begin{align*}
  h(x,t) =
  \begin{cases}
    h_l, \\
    \frac{4}{9g}\left(\sqrt{gh_l}-\frac{x-x_0}{2t}\right)^2, \\
    0,
  \end{cases}
\quad
  u(x,t) =
  \begin{cases}
    0, &\mbox{ if } x\leq x_A(t), \\
    \frac{2}{3}\left(\sqrt{gh_l}+\frac{x-x_0}{t}\right), & \mbox{ if } x_A(t) < x\leq x_B(t), \\
    0, &\mbox{ if } x_B(t) < x,
  \end{cases}
\end{align*}
with $x_A(t)=x_0 - t\sqrt{gh_l}$ and $x_B=x_0+2t\sqrt{gh_l}$.
We solve the problem up to $t=10$.
In Figures \ref{fig:rp_dry_bed_roe}-\ref{fig:rp_dry_bed_blended},
we show the solution using method \eqref{first-order_via_fluct}
with Roe's solver, Rusanov's solver and the blended solver, respectively.
The entropy glitch, which is manifested as a non-physical shock at $x=x_0$,
is present with Roe's solver. Using Rusanov's and the blended solvers
fixes the entropy glitch.
In Table \ref{table:rp_dry_bed}, we summarize the results of a convergence test.
{\color{OliveGreen}
    Due to the entropy glitch, we do not converge to the right solution with Roe solver,
    hence the drop in rate of convergence. Using any of the other solvers, we converge to the 
    correct solution. Note that using the blended solver leads to more accurate results. 
    Since the method we use is first-order accurate, we get no better than first-order convergence.}

@@

We model only the flow on the upstream side of the cylinder, since
our interest is in the resolution of the bow shock.  We take the domain
$[0,40]\times[0,100]$ with a cylinder of radius 20 centered at $(40,50)$.
Reflecting boundary conditions are imposed at the surface of the cylinder,
along with outflow conditions along the rest of the right edge of the domain.
{\color{red} The depth and velocity are set initially and (for all time) at the other
  boundaries to $h_0=1$, $u_0=5$, $v_0=0$ for a Froude number of 5. Therefore, 
  the initial condition does not contain a shock, but due to the presence of the 
  cylinder a bow shock rapidly forms and travels to its eventual steady-state position.}
  We use a $160 \times 400$ uniform Cartesian grid.
%
In Figure \ref{fig:steady_outflow}, we show results at $t=80$, after the flow has reached
a steady state.  With Roe's solver, negative values (of the water height) are generated at an early
time, leading to failure of the solver.  Therefore, we show results only for
Rusanov's, HLLEMCC by \cite{kemm2014note}, and the blended solvers.
{\color{red}
  These three methods give carbuncle free results.
  It is worth noting that the solution with Rusanov's solver is the most dissipated solution. 
  Let 
  \begin{align}\label{res}
    \xi(t^{n+1}):=\frac{||h^{n+1}-h^n||_{\ell^2}}{||h^0||_{\ell^2}}
  \end{align}
  denote the residual based on the water height. In Figure \ref{fig:residuals_cylinder_2nd}, we plot $\xi(t)$ for the
  solution obtained with each Riemann solver. With HLLEMCC or the blended solver, the residual stagnates. 
  In Figure \ref{fig:residuals_cylinder_1st} we repeat the same experiment but using the first-order method \eqref{first-order_via_fluct} instead.
  We obtain similar results with Rusanov and HLLEMCC, but now the solution with the blended solver continues to converge
  at a rate similar to that of Rusanov.  This suggests that the lack of convergence for the 2nd-order method with the
  blended solver is due to the correction terms and not to the Riemann solver itself.
  In Figure \ref{fig:cylinder_blended}, we show a closeup of the front of the shock and show the grid we use.
  Note that the grid is cartesian and that the shock is resolved across a few
  cells. Therefore, the ability of the method to deliver
  carbuncle free solutions does not depend on the shock being aligned with a grid interface. 
}

>>>>

In \cite{peery1988blunt} a numerical instability was observed to
appear near the symmetry plane in the simulation of a bow shock.
This phenomenon, subsequently dubbed ``carbuncle", has been observed by many researchers
in similar numerical experiments for the Euler equations, and many remedies
have been proposed, mainly in the form of additional numerical dissipation
\cite{quirk1997contribution,pandolfi2001numerical,dumbser2004matrix,chauvat2005shock,ismail2009proposed,shen2014stability}.
Most notably, dissipative Riemann solvers like HLLE and Rusanov suppress the carbuncle instability \cite{quirk1997contribution}.
For a recent review of numerical shock instability and work to alleviate it,
we refer to \cite[Section 2.5]{simonnumerical} and \cite{zaide2012numerical}.
The slowly-moving shock anomaly is another important unphysical behavior that might be present 
with some numerical schemes. We refer to \cite{zaide2012numerical} for a comprehensive 
discussion about this phenomenon.
In \cite{navas2019improved}, the authors focus on this problem when solving the shallow water equations.

Given the similarity of structure between the Euler equations and the shallow
water equations, it is not surprising that carbuncles appear in numerical
solutions of the latter as well \cite{kemm2014note}.
The shallow water carbuncle behaves similarly to the
Euler carbuncle; for instance, it appears when the Roe solver is used, but not when
the HLLE or Rusanov solver is used, and can also be suppressed by particular modifications
of the Roe solver \cite{kemm2014note,bader2014carbuncle}.  Another approach to avoid
shallow water carbuncles has been proposed in \cite{navas2019improved}, where it is
studied together with the slowly-moving shock anomaly and in the presence of varying bathymetry.
In this work, we propose a new Riemann solver that blends those of Rusanov and Roe.
%based on a local measure of the entropy residual, 
%as in \cite{guermond2011entropy,guermond2018second,guermond2018well}.
Since Rusanov's solver does not suffer from carbuncle instabilities,
our aim is to use it near shocks without degrading the good accuracy of Roe's solver. 
To do this, we need to find a blending criteria that can robustly and sharply locate shocks.
The entropy residual has proven to be a successful tool for this task; see for instance
\cite{guermond2011entropy,guermond2018second,guermond2018well}.
By doing this, our blended solver induces dissipation of entropy.
Entropy dissipative Riemann solvers have been used to produce carbuncle free solvers for the Euler equations;
{\color{OliveGreen} see for instance \cite{ismail2009affordable,ismail2009proposed, chizari2021cell}.}
However, to the best of our knowledge, this idea has not been used before to supress carbuncle instabilities for the shallow water equations.

Using this Riemann solver (within the second-order Lax-Wendroff-LeVeque 
finite volume scheme \cite{leveque1997wave, leveque2002finite}) supresses the formation of carbuncles 
while maintaining an accuracy similar to that of Roe solver.  
Based on our numerical experiments, using Rusanov's solver does not lead to the 
slowly-moving shock anomaly. Our blended solver inherits the same property.

@@

\subsection{Rusanov's Riemann solver}\label{sec:rusanov}
Rusanov's solver approximates the Riemann solution with a single wave traveling in each
direction.  Both waves are assumed to travel with speed $\lambda^{\max}_\imh$, which is an
upper bound on the wave speeds appearing in the true solution of the Riemann problem.
In this work, we take $\lambda_{i-1/2}^{\max}$ to be the upper bound
from \cite[Prop. 3.7]{azerad2017well}.
The waves are then given by
\begin{align}\label{rusanov_waves}
  \W_{i-1/2}^{1,\Rus}=\bfq_{i-1/2}-\bfq_{i-1}, \qquad
  \W_{i-1/2}^{2,\Rus}=\bfq_{i}-\bfq_{i-1/2},
\end{align}
where the intermediate state $\bfu_{i-1/2}$ is determined by imposing conservation:
\begin{align}\label{rus_cons}
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{1,\Rus}+\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus} = \bff(\bfq_{i})-\bff(\bfq_{i-1}).
\end{align}
The fluctuations are given by \eqref{fluct} with $s_{i-1/2}^1=-\lambda_{i-1/2}^{\max}$
and $s_{i-1/2}^2=\lambda_{i-1/2}^{\max}$, and the first- and second-order methods based
on Rusanov's solver are given by \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}, respectively.
The second-order method can also be written in flux-differencing form \eqref{flux-differencing-form}, where 
\begin{align}\label{first-order_fluxes_LLF}
  \bfF_{i+1/2} = \frac{\bff(\bfq_{i+1})+\bff(\bfq_i)}{2}
  - \frac{\lambda_{i+1/2}^{\max}}{2}(\bfq_{i+1}-\bfq_{i}).
\end{align}

\subsection{Roe's Riemann solver} \label{sec:roe}
The Roe Riemann solver is based on evaluating the flux Jacobian using Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
The waves and speeds are given by the eigenvectors and eigenvalues of the approximate
flux Jacobian obtained using these averages, resulting in
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}
and $\W_{i-1/2}^{p,\Roe}=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with
\begin{align*}
  \hat \bfr^1_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad
  \hat \bfr^2_{i-1/2} =
  \begin{bmatrix}
    0\\
    0\\
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
and the factors $\alpha^p_\imh$ obtained by solving
\begin{align}\label{system_for_alphas}
  \sum_p\W_{i-1/2}^{p,\Roe} = \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=\bfq_i-\bfq_{i-1}.
\end{align}
The first-order Roe's fluxes used in \eqref{flux-differencing-form} and \eqref{first-order_via_fluct} are given as follows:
\begin{align}\label{first-order_fluxes_Roe}
  \bfF_{i+1/2} = \frac{\bff(\bfq_{i+1})+\bff(\bfq_i)}{2} - \frac{1}{2} |\hat A_{i+1/2}|(\bfq_{i+1}-\bfq_{i}).
\end{align}

@@

@@

@@

\subsection{Dam break problem on a dry bed}\label{sec:rp_dry_bed}
We start with the one-dimensional dam break problem on a dry bed.
This problem is a canonical example that demonstrates the `entropy glitch' of Roe's solver
at transonic rarefactions. Since the baseline Roe solver we use (see \S\ref{sec:roe}) does not contain
an entropy fix, it is important to demonstrate that the blended Riemann solver fixes the entropy glitch.
We follow the setup in \cite[\S 4.1.2]{delestre2013swashes}.
The domain is given by $x\in(0,10)$, and the initial condition is ${\color{cyan}u(x,0)=0}$ and
\begin{align*}
  h(x,0)=
  \begin{cases}
    h_l & x\leq x_0=5 \\
    h_r & x >x_0,
  \end{cases}
\end{align*}
with $h_l=5\times 10^{-3}$.  In our experiments we use $h_r=1\times
10^{-15}$ to avoid division by zero.
If we use Rusanov's solver with the first-order method \eqref{first-order_via_fluct}, 
the water height is guaranteed to remain positive. Using Roe or the blended solvers 
can introduce violations of positivity. Alternatively, if we consider the second-order scheme 
\eqref{second-order_via_fluct}, the water height could become negative regardless of the Riemann solver.
To reduce the occurance of loss of positivity, we only use the first-order method 
\eqref{first-order_via_fluct} for this test problem. 
Although using Roe or the blended solvers might still lead to loss of positivity, we 
did not observe that in our experiments.
The exact solution, which can also be found in \cite{delestre2013swashes} and references therein, is
\begin{align*}
  h(x,t) =
  \begin{cases}
    h_l, \\
    \frac{4}{9g}\left(\sqrt{gh_l}-\frac{x-x_0}{2t}\right)^2, \\
    0,
  \end{cases}
\quad
  u(x,t) =
  \begin{cases}
    0, &\mbox{ if } x\leq x_A(t), \\
    \frac{2}{3}\left(\sqrt{gh_l}+\frac{x-x_0}{t}\right), & \mbox{ if } x_A(t) < x\leq x_B(t), \\
    0, &\mbox{ if } x_B(t) < x,
  \end{cases}
\end{align*}
with $x_A(t)=x_0 - t\sqrt{gh_l}$ and $x_B=x_0+2t\sqrt{gh_l}$.
We solve the problem up to $t=10$.
In Figures \ref{fig:rp_dry_bed_roe}-\ref{fig:rp_dry_bed_blended},
we show the solution using method \eqref{first-order_via_fluct}
with Roe's solver, Rusanov's solver and the blended solver, respectively.
The entropy glitch, which is manifested as a non-physical shock at $x=x_0$,
is present with Roe's solver. Using Rusanov's and the blended solvers
fixes the entropy glitch.
In Table \ref{table:rp_dry_bed}, we summarize the results of a convergence test.
Due to the entropy glitch, we do not converge to the right solution with Roe solver,
hence the drop in rate of convergence. Using any of the other solvers, we converge to the 
correct solution. Note that using the blended solver leads to more accurate results. 
Since the method we use is first-order accurate, we get no better than first-order convergence.

@@

We model only the flow on the upstream side of the cylinder, since
our interest is in the resolution of the bow shock.  We take the domain
$[0,40]\times[0,100]$ with a cylinder of radius 20 centered at $(40,50)$.
Reflecting boundary conditions are imposed at the surface of the cylinder,
along with outflow conditions along the rest of the right edge of the domain.
The depth and velocity are set initially and (for all time) at the other
boundaries to $h_0=1$, $u_0=5$, $v_0=0$ for a Froude number of 5. Therefore, 
the initial condition does not contain a shock, but due to the presence of the 
cylinder a bow shock rapidly forms and travels to its eventual steady-state position.
We use a $160 \times 400$ uniform Cartesian grid.
%
In Figure \ref{fig:cylinder_flow}, we show results at $t=80$, after the flow has reached
a steady state.  With Roe's solver, negative values (of the water height) are generated at an early
time, leading to failure of the solver.  Therefore, we show results only with the 
{\color{OliveGreen} second-order method}
\eqref{second-order_via_fluct} using Rusanov's, HLLEMCC by \cite{kemm2014note}, and the blended solvers.
These three methods give carbuncle free results.
It is worth noting that the solution with Rusanov's solver is the most dissipated solution. 
{\color{OliveGreen}
For comparison, we also show in Figure \ref{fig:cylinder_blended_1st_order} 
the solution using the first-order method \eqref{first-order_via_fluct}
with the blended solver.}
Let 
\begin{align}\label{res}
  \xi(t^{n+1}):=\frac{||h^{n+1}-h^n||_{\ell^2}}{||h^0||_{\ell^2}}
\end{align}
denote the residual based on the water height. In Figure \ref{fig:residuals_cylinder_2nd}, we plot $\xi(t)$ for the
solution obtained with each Riemann solver. With HLLEMCC or the blended solver, the residual stagnates. 
In Figure \ref{fig:residuals_cylinder_1st} we repeat the same experiment but using the first-order method \eqref{first-order_via_fluct} instead.
We obtain similar results with Rusanov and HLLEMCC, but now the solution with the blended solver continues to converge
at a rate similar to that of Rusanov.  This suggests that the lack of convergence for the 2nd-order method with the
blended solver is due to the correction terms and not to the Riemann solver itself.
{\color{OliveGreen} We have performed a similar experiment (not shown here) where we ran the simulation
up to a final time of $t=16000$ using the second-order method with the blended solver. Although, the 
residual $\xi(t)$ did not converge to zero, the simulation did not become unstable. Indeed, the 
final solution is indistinguishable from the solution in Figure \ref{fig:cylinder_blended}.}

In Figure \ref{fig:cylinder_blended}, we show a closeup of the front of the shock and show the grid we use.
Note that the grid is cartesian and that the shock is resolved across a few
cells. Therefore, the ability of the method to deliver
carbuncle free solutions does not depend on the shock being aligned with a grid interface.
